---
title: First Wave vs Second Wave
author: Arun Mitra
date: '2021-05-12'
slug: []
categories:
  - COVID
  - Epidemiology
  - R
tags:
  - COVID
publishdate: '2021-05-12T14:30:46+05:30'
comments: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

```{r load_packages_data, message=FALSE, warning=FALSE, include=FALSE}
# Load Packages
#---------------
             
## First specify the packages of interest
packages = c("tidyverse", 
             "incidence",
             "projections",
             "lubridate",
             "foreach", 
             "plotly",
             "R0",
             "magrittr",
             "EpiEstim",
             "flexdashboard",
             "shiny",
             "echarts4r",
             "distcrete",
             "epitrix",
             "furrr",
             "tictoc",
             "future",
             "data.table",
             "tibbletime",
             "mapview",
             "leaflet",
             "xts",
             "sf",
             "tbl2xts",
             "dygraphs")

## Now load or install&load all
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)

# Create Rolling Mean function
rolling_mean <- rollify(mean, window = 7)


#df <- data.table::fread("https://api.covid19india.org/csv/latest/case_time_series.csv")
#df <- data.table::fread("https://api.covid19india.org/csv/latest/districts.csv")
case_time_series <- data.table::fread("https://api.covid19india.org/csv/latest/case_time_series.csv")

state_wise <- data.table::fread("https://api.covid19india.org/csv/latest/state_wise.csv")
state_wise_backup <- state_wise
state_wise %<>% dplyr::select(State, State_code) %>% mutate(State = ifelse(State_code == "TT", "India", State))

india_time_series <- data.table::fread("https://api.covid19india.org/csv/latest/state_wise_daily.csv")
# Manipulate State_wise_time_series
india_time_series %>% 
  group_by(Status) %>% 
  mutate(DN = DN + DD) %>%
  mutate(Date = lubridate::ymd(Date_YMD)) %>% 
  dplyr::select(-c("DD", "UN", "Date_YMD")) %>% 
  ungroup() -> india_time_series

colnames(india_time_series) <- dplyr::recode(
  colnames(india_time_series), 
  !!!setNames(as.character(state_wise$State), state_wise$State_code)
)

#readr::write_rds(india_time_series, "india_time_series.rds")
```

The abrupt onset of the second wave of COVID-19 in India has taken the country by storm. With over 4,00,000 cases being reported daily, India has over 37,00,000 active cases as of 12, May 2021. This post aims to estimate the epidemiological parameters of COVID-19 epidemic in India during the first and second waves.


As discussed in the previous post, India's epidemic curve suggests an ongoing second wave of COVID-19 infections in the country. However, the magnitude and force of infection of the second wave is considerably different from the first wave. Though a multitude of factors such as the lack of a nation-wide lockdown, mutations in the virus, laxity in preventive measures and shifting priorities of the government may be attributed to the second wave. 

The epidemic curve of COVID-19 in India can be divided into two time periods based on the valley between two peaks. 


```{r epidemic_curve_division, echo=FALSE, message=FALSE, warning=FALSE}

state = "India"
#state = append(state, "India")
state_xts <- tbl_xts(india_time_series, cols_to_xts = state, spread_by = Status) 
india_xts <- tbl_xts(india_time_series, cols_to_xts = "India", spread_by = Status)

epidemic_curve_division_india <- dygraph(state_xts, 
        #main = "Weekly Average of Daily Caseload of COIVD-19 in India",
        main = paste(state)) %>%
  dyAxis("y", label = "Confirmed", logscale = TRUE) %>%
  dyAxis("y2", label = "Deceased", independentTicks = TRUE) %>%
  dySeries("Deceased", axis = 'y2') %>% 
  dyEvent("2021-02-21", "Valley between two peaks", labelLoc = "top") %>% 
  dyLegend(show = "follow") %>% 
  dyRoller(rollPeriod = 14) %>% 
  dyHighlight(highlightCircleSize = 5,
              highlightSeriesBackgroundAlpha = 0.3, 
              hideOnMouseOut = T) %>% 
  dyOptions(axisLineWidth = 1.5, fillGraph = TRUE,  drawGrid = FALSE)


epidemic_curve_division_india 

```


## Transmission Parameters

It is clear from the above epidemic curve that there were two distinct waves of COVID-19 epidemic in India. The first wave which is estimated to have peaked on 2020-09-19 with a 95% CI of being between 2020-09-18 and 2020-09-21. This has been estimated by running 100 simulations based on the epidemic curve of COVID-19 in India. The growth rate of the first wave until the peak has been 3.53% (95% CI 3.39 - 3.67) while after the peak had a rate of decline of -1.48% (95% CI -1.51 - -1.46). 

```{r first_wave, echo=FALSE, message=FALSE, warning=FALSE}
##########################################################################
# First Wave of COVID 19 Epidemic
##########################################################################

# Create Find Valley function
find_valley <- function (x, pool = TRUE) 
{
  if (!inherits(x, "incidence")) {
    stop("x is not an incidence object")
  }
  if (ncol(x$counts) > 1L && pool) {
    msg <- paste("'x' is stratified by groups", "pooling groups before finding peaks", 
                 sep = "\n")
    message(msg)
    x <- pool(x)
  }
  the_min <- apply(get_counts(x), MARGIN = 2L, FUN = which.min)
  out <- stats::setNames(x$dates[the_min], colnames(x$counts))
  out
}

# Create a function to fit the second wave
second_peak_fit <- function(x){
  x_fit <- fit(x)
  if(ind_incid_second_peak < max(ind_incid_second$dates)-7){
    date_to_split = ind_incid_second_peak
    x_fit <- fit(x, split = date_to_split)
  }
  x_fit
}

# Create Incidence Objects
options(scipen = 100)

# For India
df <- data.table::fread("https://api.covid19india.org/csv/latest/state_wise_daily.csv")
# 
# state = "AP"
# state = "KL"
state = "TT"
# state = "MH"
# state = "UP"
# state = "DN"
ind_df <- df %>%
  filter(Status == "Confirmed") %>% 
  mutate(Date = lubridate::ymd(Date_YMD)) %>% 
  dplyr::select(Date, state) %>% as_tibble() %>% 
  mutate(Cases = as.integer(rolling_mean(df %>%  filter(Status == "Confirmed") %>% pull(state)))) %>% 
  filter(!is.na(Cases))


ind_incid <- incidence::as.incidence(ind_df$Cases, ind_df$Date) 
# Epidemic Curve
#plot(ind_incid)

# Truncate the data set to one year of the first case (2021-02-15)
ind_incid_first <- ind_incid %>% subset(to = "2021-02-15")
ind_incid_second <-ind_incid %>% subset(from = "2021-02-15")

ind_incid_first_peak <- find_peak(ind_incid_first)
ind_incid_second_peak <- find_peak(ind_incid_second)

# Estimate the peak of epidemic along with 95% CI using bootstrap method
#ind_incid_first_peak_estimate <- estimate_peak(ind_incid_first)
#ind_incid_second_peak_estimate <- estimate_peak(ind_incid_second)


ind_incid_first_valley <- find_valley((ind_incid %>% subset(from = ind_incid_first_peak, to = ind_incid_second_peak)))

ind_incid_second <-ind_incid %>% subset(from = ind_incid_first_valley)

ind_first_fit <- fit(ind_incid_first, split = ind_incid_first_peak)

#plot(ind_incid_first) %>% add_incidence_fit(ind_first_fit)

ind_second_fit <- second_peak_fit(ind_incid_second)

#plot(ind_incid_second) %>% add_incidence_fit(ind_second_fit)


# Join both the plots

# Plot both the waves and estimate the optimal split
p <- ggplot(ind_df %>% filter(Date < ind_incid_first_valley)) + 
  geom_col(aes(x = Date, y = Cases), fill = "blue") + 
  geom_col(data = (ind_df %>% filter(Date >= ind_incid_first_valley)), aes(x = Date, y = Cases), fill = "red") 
p <- p %>%  add_incidence_fit(ind_first_fit) %>% add_incidence_fit(ind_second_fit)
pp <- ggplotly(p)

pp %>% layout(title = "India")


```

